Below is an example of an R code chunk which can be executed from within the R Notebook. You can do this by clicking the green triangle to the top right of the chunk area within the Notebook. If you wanted to run all the code up until, and including, this code chunk you can click the grey triangle with the green bar [^1].

[^1]: In this case there is no previous code so it would make no difference but often RMarkdown contains multiple code chunks.

```{r}
# This is an R code chunk
print("Hello cruel world")
```

So now you can handle the RMarkdown basics.  Let's move onto R.


# 2. Exploring the use of R using R {#WS2}

## 2.1 Demand for R programmers

How useful are R skills and tools? An [analysis](https://towardsdatascience.com/what-are-the-in-demand-skills-for-data-scientists-in-2020-a060b7f31b11) of 2681 data scientist job postings from 8 North American cities on January 25th, 2020 indicated the top 10 tools required by the employers are:

    Python (62%)
    SQL (40%)
    R (39%)
    Spark (21%)
    Cloud (20%)
    Amazon Web Services (20%)
    Java (19%)
    Tensorflow (16%)
    Hadoop (15%)
    SAS (13%)

The data was collected by analysing online job adverts for data scientists on the Indeed website in North America. Note that Tools is a categorical and multi-valued attribute of a job listing, i.e., one listing could have zero or more tool categories associated, thus the percentages do not sum to 100%. NB 14% of the sampled job listings are excluded (since it is stated that only 86% of adverts identified skills or tools).

**Exercise 2.1:** It's important to look at the data critically, though we must accept nothing is certain or perfect! What questions or concerns might you have?

Since we're thinking about data and data analysis, it would be helpful to visualise the data. Unfortunately, we don't have access to the raw data so we will need to reproduce it. Therefore I have 'scraped' some of the data from their article and plugged this into R (see below). Don't worry about the coding details for the present although be aware that for reasons of simplicity I'm just using Base R for the graphics (functions like `pie()` and `barplot()`). Later we will introduce the {ggplot2}[^3] package which will enable almost complete control and flexibility in producing beautiful data visualisations, however, if you are already a confident R programmer feel free to jump ahead.

```{r}
#Reproduce the data from the January 2020 required tools for Data Scientists

percent <- c(62,40,39,21,20,20,19,16,15,13)
skills <- c("Python","SQL","R","Spark","Cloud","AWS","Java","TensorFlow","Hadoop","SAS")

#Generate the pie chart

pie(percent, main = "Pie chart for most in demand language for machine learning", labels = skills, clockwise = TRUE)
?pie
```
```{r}
# NB we have already reproduced the data: percent and skill come from the 
# previous piechart chunk

# Generate a bar chart

barplot(percent, 
        main = "Bar chart of in-demand Data Scientist tools (Jan 2020)",
        sub = "Source: https://j.mp/2Ga8CD1", 
        ylab = "Percentage of listings that mention tool", 
        names.arg = skills,
        cex.names = 0.7,   # shrinks text size so fits on x-axis
        col="darkolivegreen", 
        density = c(100, 50, 100, 50, 50, 50, 50, 50, 50, 50))
```
In the previous task you may well have largely relied on the provided R code. Now we move onto learning about basic R syntax and some coding exercises for you to undertake.

Suppose I'm trying to calculate how much money I'm going to make from my *Modern Data* book. There are several different things I will need to store:

-   `sales` how many copies I'll sell `sales`\
-   `price` the price per copy
-   `numberStudents` number of students in the class (assume I sell one copy per student)
-   `recommend` the proportion of students so impressed that they recommend the book to friends and family
-   `income` the income generated

Let's turn this into R.

```{r}
#Compute income from sales of Ama's book

price <- 10 
numOfStudents <- 100
recommend <- 0.5

sales <- numOfStudents * (1 + recommend)
income <- sales * price
paste("Ama would earn", income)
```
Although we could just output `income`, I have added some text to make the output of the code easier to interpret with the `paste()` function to combine a character string literal delimited by quotation marks and the calculated income.

**Exercise 3.1:** Change the number of students to 200 and the recommendation rate to 0.2. What's my new estimated income?

**Exercise 3.2:** Update the sales income R code to handle tax. Assume that income is taxed at a flat rate of 20%. Modify the above R code to output gross income and net (after tax) income. Using RStudio you can edit this Week 1 Worksheet Notebook.

```{r}
numOfStudents <- 200

recommend <- 0.2

income <- numOfStudents * (1 + recommend)
print(income)

grossIncome <- income

netIncome <- grossIncome * (1 - 0.2)

print(netIncome)
```
HINT: I recommend you define a new numeric variable to store the tax rate and another for net income. For clarity you might want to rename `income` to make it obvious that it stores gross income.

**Extension Exercise 3.3:** How would you answer the question: it's possible to write a much shorter R program without any extra variables, so why bother?

It's very *important* to get into the habit of writing easy to follow R code. For this reason I have added comments and tried to select meaningful variable names.

## 3.2 Using vectors

Recall that at a Vector is a repeated group of the **same** data type, e.g., a vector of character strings.

```{r}
#Define a new vector

greetings <- c("Hi!", "Hello", "Good morning", "Saludo", "Hej")
greetings 

```
We can also check the type of `greetings`

```{r}
is.vector(greetings)
```

**Exercise 3.4:** Check the type of the variable `price` (from Section 3.1). What output do you expect?

To fully exploit the power of vectors we can access subsets of elements. To do this we use the square bracket notation `[<n>]` where $n$ is the position of the required vector element.


```{r}
#Different ways to index vector elements
greetings[2]

n <- 3

greetings[n]
greetings[1:n]
```

Notice in the above R chunk, that we can access more than one element at a time by using the `:` operator to return a range of elements. The range can be specified by literals as in `[1:3]` or integer[^4] variables e.g., `[m:n]`.

[^4]: If the index variables are not integers R will do its best to coerce the values to integers.

## 3.3 Using data frames

Data frames are amongst the most useful of the different data structures. These are two dimensional and unlike 2-dimensional matrices can comprise different constituent data types.

Let's start by examining a built-in data set mt-cars that stores information about motor cars from the 1970s that is stored as a data frame.

**Exercise 3.5:** How many variables and observations does `df` have? HINT: There are functions `nrow()` and `ncol()` if you can't be bothered to count.

You can use the `$` operator to specify a column/variable from a data frame, for instance:

Let's start by examining a built-in data set mt-cars that stores information about motor cars from the 1970s that is stored as a data frame.

```{r}
# Assign the built in data object mtcars to a new data frame df
df <- mtcars

# View the data frame in a new window in RStudio
View(df)
```

**Exercise 3.5:** How many variables and observations does `df` have? HINT: There are functions `nrow()` and `ncol()` if you can't be bothered to count.

You can use the `$` operator to specify a column/variable from a data frame, for instance:

```{r}
# Display the observations for cylinder numbers from df
df$cyl

# Apply the median function
median(df$cyl)
```
**Exercise 3.6:** Using the `$` operator determine the mean and minimum mpg for cars in df.

**Exercise 3.7:** Update the number of gears for the Mazda RX4 from 4 to 6. HINT: `mt$gear` is a vector so you can use an index to reference a particular element.

So we can use the dollar operator to identify columns/variables. What about rows/observations? There are several ways to achieve this. Using indices via the `[ ]` notation notice that since a data frame has two dimensions two arguments are needed, but since we want the entire row we can leave the argument after the comma blank.

```{r}
# To retrieve the 2nd row we could have.
df[2,]
```

**Exercise 3.8:** Modify the above code to retrieve columns 1-3. HINT: remember the `:` operators from when we were manipulating vectors.

The other way to retrieve specific observations is when the rows have names e.g., "Mazda RX4", as in this case.

```{r}
# Retrieve the observation named "Valiant"
df["Valiant",]
```

